Compositional Certified Resource Bounds\ifshort{\vspace{-.15cm}}{}

### 2014-11-10 paper/main.tex
freshen section 3, update plan

% Structure:
%
% 1) Intro
% 2) Overview and Examples
% 3) Cost Semantics for Clight
% 4) Compositional Resource-Bound Analysis
% 5) Automatic Inference via LP Solving
% 6) Auxiliary Variables and User Interaction
% 7) Quantitative Hoare Logic and Soundness Proof
% 8) Experimental Evaluation
% 9) Related Work
% 10) Conclusion

@@

\sectskip
\section{Automatic Amortized Analysis}
\label{sec:AAA}
\aftersectskip

In this section we present the automatic amortized analysis that we
use in our implementation to derive resource bounds for C
programs.

@@

To find derivations in the quantitative Hoare logic automatically, we
have to focus on bounds that have a certain shape.  The general form
of \emph{potential functions} (or assertions) that we consider is
$$
  \Phi(\state) = q_0 + \sum_{x, y \in \dom\state \land x \neq y}
    q_{(x, y)} \cdot |\inter {\state(x)} {\state(y)}| \, .
$$
%
Here $\state : (\text{Locals} \to \Z) \times (\text{Globals} \to \Z)$
is again a simplified program state as introduced in \pref{sec:sem},
$|\inter a b| = max(0,b-a)$, and $q_i \in \Qplusz$.  To simplify the
references to the linear coefficients $q_i$, we introduce an \emph{index
  set} $I$.  This set is defined to be $\{0\} \cup \{(x, y) \mid x, y
\in \Var \land x \neq y \}$.  Each index $i$ corresponds to a \emph{base function}
$f_i$ in the potential function: $0$ corresponds to the constant
function $\state \mapsto 1$, and $(x,y)$ corresponds to $\state \mapsto
|\inter {\state(x)} {\state(y)}|$.  Using these notations we can
rewrite the above equality as\iffull{
$$
  \Phi = \sum_{i \in I} q_i f_i.
$$}
{ $\Phi = \sum_{i \in I} q_i f_i$.}  We often write $xy$ to denote the
index $(x,y)$.  \iffull{The family $(f_i)_{i\in I}$ is actually a basis (in
the linear-algebra sense) for potential functions.}{}  This allows us to
uniquely represent any linear potential function
$\Phi$ as a \emph{quantitative annotation} $Q = (q_i)_{i \in I}$, that
is, a family of \ifshort{non-negative}{} rational numbers where only
a finite number of elements are not zero. \iffull{ For soundness, we
  require that potential functions always give non-negative results.
  This is acheived by restricting coefficients in a quantitative
  annotation to be always non-negative.}{}

@@

\paragraph{Logical State.}

In addition to the quantitative annotations our automatic amortized
analysis needs to maintain a minimal logical state to justify certain
operations on quantitative annotations.  For example when analyzing
the code $x \gets x + y$, it is helpful to know the sign of $y$ to
determine which intervals will increase or decrease.  The knowledge
needed by our rules can be inferred by local reasoning (i.e., in
basic blocks without recursion and loops) within usual theories
(e.g. Presburger arithmetic or bit vectors).

In contrast to the quantitative annotations, logical contexts $\Gamma$
in the presented inference system are left abstract.  This allows for
simpler rules and leaves room for future improvements.
%
Our implementation uses conjunctions of linear inequalities as
logical state.  We never compute fixpoints and take $\top$
as pre- and postconditions for functions. It has proved to be
sufficient for the variety of examples we are interested in or found
in the literature.

@@

@@

\end{mathpar}
\caption{\iffull{Inference rules}{Selected inference rules} of the quantitative analysis.}
\label{fig:auto}
\end{figure*}
}{}

@@

\end{mathpar}
\caption{\iffull{Inference rules}{Selected inference rules} of the quantitative analysis.}
\label{fig:auto}
\end{figure*}
}{}

@@

The inference system for the automatic amortized analysis is defined
in~\pref{fig:auto}.  The inference rules derive judgements of the form
$$
  (\Gamma_B, Q_B); (\Gamma_R, Q_R);
    (\Gamma, Q) \vdash S \dashv (\Gamma', Q').
$$
These judgements correspond to the logic judgements $\Delta; B; R
\vdash \htriple P S Q$ where each assertion splits in two parts, a
logical part $\Gamma$ and a quantitative part $Q$.  That means that
$(\Gamma_B, Q_B)$ is the postcondition of \code{break} statements,
$(\Gamma_R, Q_R)$ is the postcondition for return statements, and
$(\Gamma, Q) \vdash S \dashv (\Gamma', Q')$ can be understood as a
simple Hoare triple.  We leave out the function context $\Delta$ in
the presentation to avoid cluttering notations.  Instead we assume a
fixed function context $\Delta$ that is implicit on all judgements.
Function contexts are treated exactly as in the logic.  % We call
% precondition the pair $(\Gamma, Q)$ and postcondition the pair
% $(\Gamma', Q')$.
The correspondence between judgements of the
automation and the ones of the quantitative logic is made formal by
our soundness proof of the automatic amortized analysis at the end of
this section.

As a convention, if $Q$ and $Q'$ are quantitative annotations then we
assume that $Q = (q_i)_{i\in I}$ is a family with elements $q_i$, $Q'
= (q'_i)_{i \in I}$, etc.  The notation $Q \pm n$ used in many rules
defines a new context $Q'$ such that $q'_0 = q_0 \pm n$ and $\forall i
\neq 0 .\, q'_i = q_i$.  We have the implicit side condition that all
rational coefficients are non-negative.  Finally, if a rule mentions
$Q$ and $Q'$ and leaves the latter undefined at some index $i$ we will
implicitly assume that $q'_i = q_i$.

We describe the automatic amortized analysis for a subset of
expressions of Clight.  Assignments must have the form $x \gets
y$ or $x \gets x \pm y$.  In the implementation, a Clight
program is converted into this form prior to analysis without changing
the resource cost.  This achieved by using a series of \emph{cost-free
  assignments} that do not result in additional cost in the
analysis.  Non-linear operations such as $x \gets z*y$ are simply
handled by assigning zero potential to coefficients like $q_{xa}$ and
$q_{ax}$ that contain $x$ after the assignment.

@@

Most of the rules correspond exactly to the respective rules of the
quantitative Hoare logic.
\iffull{
  For example, the rule {\sc Q:Skip} simply reuses its postcondition
  as precondition.  This is sound since evaluating a $\code{skip}$
  statement never costs any resources and does not change the program
  state.  Similarly, the rule {\sc Q:Tick} enforces that the
  quantitative annotation of the precondition is at least $M_t(n)$ to
  pay for the resource consumption.  Since the program state is also
  unchanged in this case, the logical part $\Gamma$ of the
  precondition can soundly be reused in the postcondition.  In the
  rule {\sc Q:Break}, where the linear control flow is broken, the
  postcondition $(\Gamma', Q')$ can be arbitrary.  The precondition is
  taken from the break part of the judgement similarly as it is done
  in the quantitative logic.
}
%
{ Examples include {\sc Q:Skip, Q:Tick, Q:Break, Q:Seq, Q:Assert}, and
  {\sc Q:Loop}.  We only show {\sc Q:Seq} and {\sc Q:Loop}.  The other
  rules can be found in the extended version.
}
%
The rule {\sc Q:Return} exhibits a slight
difference to the quantitative logic.  While we have a function $\Z
\to \Assn$ to represent return values in the logic, we assume that a
special variable $\ret$ is part of the index set of $Q_R$.  The
substitution $Q[\ret/x]$ denotes an potential annotation $Q'$ with
$q'_{xy} = q_{\ret y}$ and $q'_{yx} = q_{y \ret}$ for all $y$ and
$q_i'=q_i$ otherwise.

Most interesting are the rules {\sc Q:IncP}, {\sc Q:DecP}, and {\sc
  Q:Inc} for increments and decrements, which describe how the
potential is distributed after a size change of a variable.  The rule
{\sc Q:IncP} is for increments $x \gets x + y$ and {\sc Q:DecP} is
for decrements $x \gets x - y$, they both apply only when we can deduce
from the logical context $\Gamma$ that $y \ge 0$.
Of course, we have symmetrical rules {\sc
  Q:IncN} and {\sc Q:DecN} that can be applied if $y$ is not positive.
The rules are equivalent in the case where $y=0$.  Finally, the rule
{\sc Q:Inc} can be applied if we cannot deduce any information about
$y$.  In the implementation, the rules for increments and decrements
are combined into one syntax directed rule where the specialized rules
take priority over {\sc Q:Inc}.

@@

The rule {\sc Q:Weak} is not syntax directed.  In the implementation
we apply {\sc Q:Weak} before loops and between the two statements of a
sequential composition.  \iffull{We could integrate weakening into every
syntax directed rule but this simple heuristic helps to make the
analysis efficient.}{}  The high-level idea of {\sc Q:Weak} is the same
as in the rule {\sc L:Weak} of the quantitative logic: If we have a
sound judgement, then it is sound to add more potential to the
precondition and remove potential from the postcondition.  The concept
of \emph{more potential} is formalized by the relation $Q'
\succeq_\Gamma Q$ that is defined in the rule {\sc Relax}.  Here, $Q$
and $Q'$ are potential annotations and $\Gamma$ is a logical context.
The rule {\sc Relax} deals also with the important task of transferring
constant potential (represented by $q_0$) to interval sizes and vice
versa.  If we can deduce from the logical context that the interval
size $|\inter x y| \geq l$ is larger then a constant $l$ then we can
transfer potential $q_{xy} {\cdot} |\inter x y|$ form the interval to
constant potential $l{\cdot}q_{xy}$ and guarantee that we do not gain
potential.  Conversely, if $|\inter x y| \leq u$ for a constant $u$
then we can transfer constant potential $u{\cdot}q_{xy}$ to the
interval potential $q_{xy} {\cdot} |\inter x y|$ without gaining
potential.

\paragraph{Automatic Inference via LP Solving.}

@@

\paragraph{Soundness Proof.}

>>>>

% Structure:
%
%    Abstract                              1c
% 1) Intro                                 3c
% 2) The Potential Method                  1c
% 3) Compositional Resource-Bound Analysis 4c
%    - Gulwani/ examples
% 4) Derivation Rules                      2c
% 5) Automatic Inference via LP Solving    2c
% 6) Logical State and User Interaction    2c
% 7) Soundness Proof                       2c
%    1) Cost Semantics for Clight
%    2) Quantitative Hoare Logic
% 8) Experimental Evaluation               2c
% 9) Related Work                          1c
% a) Conclusion                            1c

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\sectskip
\section{Automated Reasoning with \toolname}
\label{sec:AAA}
\aftersectskip

In this section we present formally the analysis that we
use in our implementation to derive resource bounds for C code.

@@

To find resource bounds automatically, we first need to restrict our
search space.  This is done by focusing on bounds that have a certain
shape.  The general form of \emph{potential functions} that we consider is
$$
  \Phi(\state) = q_0 + \sum_{x, y \in \dom\state \land x \neq y}
    q_{(x, y)} \cdot |\inter {\state(x)} {\state(y)}| \, .
$$
%
Here $\state : (\text{Locals} \to \Z) \times (\text{Globals} \to \Z)$
is a simplified program state that maps variable names to integers,
$|\inter a b| = \max(0,b-a)$, and $q_i \in \Qplusz$.  To simplify the
references to the linear coefficients $q_i$, we introduce an \emph{index
  set} $I$.  This set is defined to be $\{0\} \cup \{(x, y) \mid x, y
\in \Var \land x \neq y \}$.  Each index $i$ corresponds to a \emph{base function}
$f_i$ in the potential function: $0$ corresponds to the constant
function $\state \mapsto 1$, and $(x,y)$ corresponds to $\state \mapsto
|\inter {\state(x)} {\state(y)}|$.  Using these notations we can
rewrite the above equality as\iffull{
$$ \Phi = \sum_{i \in I} q_i f_i.$$}{
$\Phi = \sum_{i \in I} q_i f_i$.}%
We often write $xy$ to denote the
index $(x,y)$.  \iffull{The family $(f_i)_{i\in I}$ is actually a basis (in
the linear-algebra sense) for potential functions.}{}  This allows us to
uniquely represent any linear potential function
$\Phi$ as a \emph{quantitative annotation} $Q = (q_i)_{i \in I}$, that
is, a family of \ifshort{non-negative}{} rational numbers where only
a finite number of elements are not zero. \iffull{ For soundness, we
  require that potential functions always give non-negative results.
  This is acheived by restricting coefficients in a quantitative
  annotation to be always non-negative.}{}

@@

\paragraph{Abstract Program State.}

In addition to the quantitative annotations our automatic amortized
analysis needs to maintain a minimal abstract state to justify certain
operations on quantitative annotations.  For example when analyzing
the code $x \gets x + y$, it is helpful to know the sign of $y$ to
determine which intervals will increase or decrease.  The knowledge
needed by our rules can be inferred by local reasoning (i.e. in
basic blocks without recursion and loops) within usual theories
(e.g. Presburger arithmetic or bit vectors).

In contrast to the quantitative annotations, logical contexts $\Gamma$
in the presented inference system are left abstract.  This allows for
simpler rules and leaves room for future improvements.
Our implementation finds these logical contexts using abstract
interpretation using the domain of linear inequalities.  We observed that
the rules of the analysis often requires only minimal local knowledge.
This means that it is not necessary for us to compute precise loop
invariants and only a rough fixpoint (e.g. keeping only inequalities on
variables unchanged by the loop) is sufficient to obtain good bounds.

@@

{ \!\! B; R; (\Gamma, Q) \vdash \code{skip} \dashv (\Gamma, Q) \!\!}

@@

\end{mathpar}
\caption{Inference rules of the quantitative analysis.}
\label{fig:auto}
\end{figure*}

@@

\end{mathpar}
\caption{Selected inference rules of the quantitative analysis.}
\label{fig:auto}
\end{figure*}
}{}

@@

The inference system for the automatic amortized analysis is defined
in~\pref{fig:auto}.  The inference rules derive judgements of the form
$$
  (\Gamma_B, Q_B); (\Gamma_R, Q_R);
    (\Gamma, Q) \vdash S \dashv (\Gamma', Q').
$$

These judgements are similar to classic Hoare triples in the presence
of the early exit statements \code{break} and \code{return}.  The part
$(\Gamma_B, Q_B)$ is the postcondition used when breaking out of
a loop.  For example, it is set upon entry of looping statements by
the rule {\sc Q:Loop} and used before \code{break} statements by the
rule {\sc Q:Break}.  Similarly $(\Gamma_R, Q_R)$ is the postcondition
used before returning from a function call.  The part $(\Gamma, Q) \vdash
S \dashv (\Gamma', Q')$ of the judgement can be seen as a regular
Hoare triple.  All assertions are split into two parts, the logical
part and the quantitative part.  The quantitative part $Q$ represents a
potential function as a collection of non-negative numbers $q_i$ indexed
by the index set $I$.  The logical part $\Gamma$ is left abstract but
is enforced by our derivation system to respect classic Hoare logic
constraints.

% These judgements correspond to the logic judgements $\Delta; B; R
% \vdash \htriple P S Q$ where each assertion splits in two parts, a
% logical part $\Gamma$ and a quantitative part $Q$.  That means that
% $(\Gamma_B, Q_B)$ is the postcondition of \code{break} statements,
% $(\Gamma_R, Q_R)$ is the postcondition for return statements, and
% $(\Gamma, Q) \vdash S \dashv (\Gamma', Q')$ can be understood as a
% simple Hoare triple.  We leave out the function context $\Delta$ in
% the presentation to avoid cluttering notations.  Instead we assume a
% fixed function context $\Delta$ that is implicit on all judgements.
% Function contexts are treated exactly as in the logic.  % We call
% % precondition the pair $(\Gamma, Q)$ and postcondition the pair
% % $(\Gamma', Q')$.
% The correspondence between judgements of the
% automation and the ones of the quantitative logic is made formal by
% our soundness proof of the automatic amortized analysis at the end of
% this section.

As a convention, if $Q$ and $Q'$ are quantitative annotations
we assume that $Q = (q_i)_{i\in I}$ and $Q' = (q'_i)_{i \in I}$.
The notation $Q \pm n$ used in many rules defines a new context $Q'$
such that $q'_0 = q_0 \pm n$ and $\forall i \neq 0 .\, q'_i = q_i$.
In all the rules, we have the implicit side condition that all rational
coefficients are non-negative.  Finally, if a rule mentions $Q$ and $Q'$
and leaves the latter undefined at some index $i$ we will implicitly
assume that $q'_i = q_i$.

We describe the automatic amortized analysis for a subset of
expressions of Clight.  Assignments are restricted to the form $x \gets
y$ or $x \gets x \pm y$.  In the implementation, a Clight
program is converted into this form prior to analysis without changing
the resource cost.  This is achieved by using a series of \emph{cost-free
  assignments} that do not result in additional cost in the
semantics.  Non-linear operations such as $x \gets z*y$ or $x \gets
a[y]$ are simply handled by assigning zero potential to coefficients
like $q_{xa}$ and $q_{ax}$ that contain $x$ after the assignment.

@@

The rules of our derivation system must serve two purposes.  They must
\emph{attach} potential to certain program variable intervals and use
this potential, when it is allowed, to \emph{pay} for resource consuming
operations.  These two purposes are easily illustrated on the {\sc Q:Skip}
rule.  This rule reuses its precondition as postcondition.  This is
explained by two facts: first, no resource is consumed by the skip
operation, thus no potential needs to be taken out of the precondition
to pay for the evaluation; second, the program state is not changed by
the execution of a skip statement, thus all the potential available
before the execution of the skip statement is still available after.
These two observations would also justify a rule with the side-condition
$Q' \le Q$, because some potential is always safely lost.  For the sake
of orthogonality, it is not how we present the rule and this form of
weakening is in fact expressed as the stand-alone rule {\sc Q:Weak}.

% Most of the rules correspond exactly to the respective rules of the
% quantitative Hoare logic.
% \iffull{
%   For example, the rule {\sc Q:Skip} simply reuses its postcondition
%   as precondition.  This is sound since evaluating a $\code{skip}$
%   statement never costs any resources and does not change the program
%   state.  Similarly, the rule {\sc Q:Tick} enforces that the
%   quantitative annotation of the precondition is at least $M_t(n)$ to
%   pay for the resource consumption.  Since the program state is also
%   unchanged in this case, the logical part $\Gamma$ of the
%   precondition can soundly be reused in the postcondition.  In the
%   rule {\sc Q:Break}, where the linear control flow is broken, the
%   postcondition $(\Gamma', Q')$ can be arbitrary.  The precondition is
%   taken from the break part of the judgement similarly as it is done
%   in the quantitative logic.
% }
% %
% { Examples include {\sc Q:Skip, Q:Tick, Q:Break, Q:Seq, Q:Assert}, and
%   {\sc Q:Loop}.  We only show {\sc Q:Seq} and {\sc Q:Loop}.  The other
%   rules can be found in the extended version.
% }
% %
% The rule {\sc Q:Return} exhibits a slight
% difference to the quantitative logic.  While we have a function $\Z
% \to \Assn$ to represent return values in the logic, we assume that a
% special variable $\ret$ is part of the index set of $Q_R$.  The
% substitution $Q[\ret/x]$ denotes an potential annotation $Q'$ with
% $q'_{xy} = q_{\ret y}$ and $q'_{yx} = q_{y \ret}$ for all $y$ and
% $q_i'=q_i$ otherwise.

Most interesting are the rules {\sc Q:IncP}, {\sc Q:DecP}, and {\sc Q:Inc}
which describe how the potential is distributed after a size change of
a variable.  The rule {\sc Q:IncP} is for increments $x \gets x + y$
and {\sc Q:DecP} is for decrements $x \gets x - y$, they both apply only
when we can deduce from the logical context $\Gamma$ that $y \ge 0$.
Of course, we have symmetrical rules {\sc Q:IncN} and {\sc Q:DecN}
that can be applied if $y$ is not positive.  The rules are equivalent
in the case where $y=0$.  Finally, the rule {\sc Q:Inc} can be applied
if we cannot deduce any information about $y$.  In the implementation,
the rules for increments and decrements are combined into one syntax
directed rule where the specialized rules take priority over {\sc Q:Inc}.

@@

The rule {\sc Q:Weak} is not syntax directed.  In the implementation
we apply {\sc Q:Weak} before loops and between the two statements of a
sequential composition.  \iffull{We could integrate weakening into every
syntax directed rule but this simple heuristic helps to make the analysis
efficient.}{}  The high-level idea of {\sc Q:Weak} is the following: If
we have a sound judgement, then it is sound to add more potential to the
precondition and remove potential from the postcondition.  The concept of
\emph{more potential} is formalized by the relation $Q' \succeq_\Gamma Q$
that is defined in the rule {\sc Relax}.  Here, $Q$ and $Q'$ are potential
annotations and $\Gamma$ is a logical context.  The rule {\sc Relax}
deals also with the important task of transferring constant potential
(represented by $q_0$) to interval sizes and vice versa.  If we can deduce
from the logical context that the interval size $|\inter x y| \geq l$ is
larger then a constant $l$ then we can transfer potential $q_{xy} {\cdot}
|\inter x y|$ form the interval to constant potential $l{\cdot}q_{xy}$
and guarantee that we do not gain potential.  Conversely, if $|\inter
x y| \leq u$ for a constant $u$ then we can transfer constant potential
$u{\cdot}q_{xy}$ to the interval potential $q_{xy} {\cdot} |\inter x y|$
without gaining potential.

\sectskip
\section{Automatic Inference via LP Solving.}
\label{sec:lp}
\aftersectskip

@@

\sectskip
\section{Soundness Proof}
\label{sec:soundness}
\aftersectskip
